\chapter{Summary}

The closer we come to human technological advancement, the more people start to get involved in this splendid world of programming. At the same time, we live in the Renaissance of programming languages. The importance of static analysis tools is to use correctly the programming languages, according to the guidelines and standards, to make it more professional. This thesis aimed to the part of this community effort.


It presented an overview of LLVM and clang-tidy in particular, and a solution for static analysis check of a particular set of bugs in C/C++ which are caused by unexpected behavior of the program. The thesis concentrated on explaining to the reader the basic examples of the consequences of violating those warnings, and a short guideline into Clang AST and marcher creation pipeline.  In the course of the development, two possible solutions for the matcher were found, and both of those were tested and checked through the categories in order to find the best match for the community code. 

Even though the top-down matcher was chosen in the end, the important lessons were taken from the bottom-up solution as well. There is still a big field of performance testing and adjustments which can make the checker work faster, which has a big importance for the IDE in particular. Moreover, a vast field of code readability and its implementations was found, which has a lot of good theory material has not so many practice references.

Altogether, it was a great experience working in the compiler front-end field, both as getting practice knowledge and theory application, plus contributing to the largest codebase in the world. Therefore, it is a great study material for the demonstration of the knowledge and experience achieved through my bachelor studies, and my personal triumph over one more field of computer science, and life in general.  